---
title: "Analysis for 3D image"
author: "Xiyu Peng"
date: 2025-07-15
output: html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

NOTE: This tutorial is only for the development version of SpaTopic. Please install the latest version of SpaTopic from Github. It is not included in v1.2.0 on CRAN

```{r, eval=FALSE}
devtools::install_github("xiyupeng/SpaTopic")
```

This is still under active development. We welcome any feedbacks.

## Set up

```{r package}
library(SpaTopic)
packageVersion("SpaTopic")
library(sf)
```

Again, we will work on a dataset of 9 spleen tissue images. We will focus on one image and illustrate how to use `SpaTopic` on a 3D image dataset.
The dataset object can be downloaded from [here](https://drive.google.com/drive/folders/1_mJUjzQXWgUZlwUaLq0HKxX-aqgiQ8eD?usp=sharing), with original source from a [Cell paper](https://www.cell.com/cell/fulltext/S0092-8674(18)30904-8?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0092867418309048%3Fshowall%3Dtrue). 

This dataset includes nine images: three control normal BALBc spleens (BALBc 1-3) and six MRL/lpr spleens (samples 4-9) at varying disease stages—early (MRL 4-6), intermediate (MRL 7-8), and late (MRL 9) (Figure 5A). 
Using a 30-plex protein marker panel, the study identified 27 major splenic-resident cell types across the nine tissue images.
We use the cell type annotation in the original paper.

```{r setup}
x<-readRDS(file = "~/Documents/Research/github/SpaTopic_benchmarking/data/mouse_spleen_data/spleen_df.rds")
names(x)
```

We firstly prepare this dataset as the required input for SpaTopic. 
For multiple images, SpaTopic requires a list of data frames as input, each with columns

- image
- X
- Y
- Z (additional axis for 3D image)
- type

Z should be the axis with the **lowest** range (max(Z) - min(Z)).

```{r}
select_column<-function(data){
  data_select<-as.data.frame(cbind(data$sample, data$sample.X, data$sample.Y, data$sample.Z, data$cluster))
  colnames(data_select)<-c("image","X","Y","Z","type") # we also save the coordinate on Z axis
  data_select$image<-gsub("-", ".", data_select$image)
  data_select
}
dataset<-lapply(x,select_column)
names(dataset) ## names of the 9 images
head(dataset[["BALBc-1"]]) ## df for the first image
```

## Running SpaTopic on 3D image

SpaTopic now supports full 3D spatial analysis, allowing you to leverage the complete spatial information in your tissue images. 
To enable 3D analysis, simply set `axis = "3D"` in the `SpaTopic_inference` function. 


**How it works?** The package considers spatial relationships between cells in all three dimensions (X, Y, and Z), where Z represents the depth axis. 
For thick tissue sections, we implemented an `adaptive` sampling strategy that selects anchor cells across Z planes. 
The sampling density along the Z axis is controlled by the `z_cellsize` parameter, which we recommend setting to twice the `region_radius` by default. 

Note that the Z axis should be the dimension with the smallest range (max(Z) - min(Z)) in your coordinate system.

For this image, we set ntopics = 4, similar as in the 
[paper](https://www.cell.com/cell/fulltext/S0092-8674(18)30904-8?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0092867418309048%3Fshowall%3Dtrue) with four domains:

- B-zone, marginal zone, PALS, red pulp 

 
```{r}
image1<-dataset[["BALBc-1"]]
system.time(gibbs.res<-SpaTopic_inference(image1, ntopics = 4, sigma = 20, region_radius = 150, burnin = 1500, axis = "3D"))
```


```{r,warning=FALSE,message = FALSE}
spleen_objects<-list()
library(Seurat)
for(i in names(x)){
  counts<-as.matrix(x[[i]][,c(2:30,34)])
  rownames(counts)<-1:nrow(counts)  ### to avoid bug for mismatch
  subset<-CreateSeuratObject(counts = t(counts), assay = "protein")
  subset$X<-x[[i]]$sample.X
  subset$Y<-x[[i]]$sample.Y
  subset$cluster<-x[[i]]$cluster
  subset$sample<-x[[i]]$sample
  coords <- CreateFOV(
    data.frame(X = subset$X,Y = subset$Y),
    type = c("centroids"),assay = "protein"
  )
  i<-gsub("-", ".", i) # seurat package does not like -
  subset[[i]] <- coords
  spleen_objects[[i]]<-subset
}
```


```{r}
spleen_objects[["BALBc.1"]]$Topic_3D<-as.factor(gibbs.res$cell_topics)
ImageDimPlot(spleen_objects[["BALBc.1"]], fov = "BALBc.1", group.by = "Topic_3D", axes = FALSE, dark.background = F)
```

## Visualize result in 3D

With the help with R package `rgl`, we are able to visualize the topic distribution in 3D.

```{r}
library(rgl)
palatte<- c("#F8766D","#7CAE00","#00BFC4","#C77CFF")
names(palatte)<-as.character(1:4)

image_data<-dataset[["BALBc-1"]]
image_data$color <- palatte[spleen_objects[["BALBc.1"]]$Topic_3D]

plot3d( 
  x=image_data$X, y=image_data$Y, z=image_data$Z, 
  col = image_data$color, 
  xlab="X", ylab="Y", zlab="Z")

rglwidget()
```


