---
title: "Analysis across Multiple Images"
author: "Xiyu Peng"
date: 2025-04-17
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

In this tutorial, we will show how to apply SpaTopic to jointly identify topics across multiple images.

## Set-up

We will work on 9 spleen tissue images and illustrate how to apply SpaTopic to multiple images to jointly identify topics. The dataset object can be downloaded from [here](https://drive.google.com/drive/folders/1_mJUjzQXWgUZlwUaLq0HKxX-aqgiQ8eD?usp=sharing), with original source from a [Cell paper](https://www.cell.com/cell/fulltext/S0092-8674(18)30904-8?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS0092867418309048%3Fshowall%3Dtrue). 
This dataset includes nine images: three control normal BALBc spleens (BALBc 1-3) and six MRL/lpr spleens (samples 4-9) at varying disease stages—early (MRL 4-6), intermediate (MRL 7-8), and late (MRL 9) (Figure 5A). 
Using a 30-plex protein marker panel, the study identified 27 major splenic-resident cell types across the nine tissue images.
We use the cell type annotation in the original paper.

```{r setup}
x<-readRDS(file = "~/Documents/Research/github/SpaTopic_benchmarking/data/mouse_spleen_data/spleen_df.rds")
names(x)
```

We firstly prepare this dataset as the required input for SpaTopic. 
For multiple images, SpaTopic requires a list of data frames as input, each with columns

- image
- X
- Y
- type

You may add additional axis Y2 for 3D image.

```{r}
select_column<-function(data){
  data_select<-as.data.frame(cbind(data$sample, data$sample.X, data$sample.Y, data$sample.Z, data$cluster))
  colnames(data_select)<-c("image","X","Y","Y2","type") # we also save the coordinate on Z axis
  data_select$image<-gsub("-", ".", data_select$image)
  data_select
}
dataset<-lapply(x,select_column)
names(dataset) ## names of the 9 images
head(dataset[["BALBc-1"]]) ## df for the first image
```

## Run SpaTopic

Once we have prepared data, we can start to run SpaTopic with number of topic = 6. The initialization may takes time, and you may run the algorithm in parallel `do.parallel = TRUE` with multiple cores `n.cores = 8`.

```{r,eval=FALSE}
library(SpaTopic)
library(sf)
## NOT RUN, TAKES ABOUT 10~15 MIN across 9 images with single core.
system.time(gibbs.res<-SpaTopic_inference(dataset, ntopics = 6, sigma = 20, region_radius = 150, burnin = 1500))
save(gibbs.res,file = "~/Documents/Research/github/SpaTopic_benchmarking/data/mouse_spleen_data/gibbs.res.v2.rdata")
```

```{r}
load(file = "~/Documents/Research/github/SpaTopic_benchmarking/data/mouse_spleen_data/gibbs.res.v2.rdata")
```

## Visualize the result

We will still use Seurat Package to help for visualization.
First, let's create a list of Seurat Object.

```{r,warning=FALSE,message = FALSE}
spleen_objects<-list()
library(Seurat)
for(i in names(x)){
  counts<-as.matrix(x[[i]][,c(2:30,34)])
  subset<-CreateSeuratObject(counts = t(counts), assay = "protein")
  subset$X<-x[[i]]$sample.X
  subset$Y<-x[[i]]$sample.Y
  subset$cluster<-x[[i]]$cluster
  subset$sample<-x[[i]]$sample
  coords <- CreateFOV(
    data.frame(X = subset$X,Y = subset$Y),
    type = c("centroids"),assay = "protein"
  )
  i<-gsub("-", ".", i) # seurat package does not like -
  subset[[i]] <- coords
  spleen_objects[[i]]<-subset
}
```

We first plot the spatial distribution of 27 cell types across 9 images.

```{r celltype, fig.width=12, fig.height=12}
library(ggplot2)
library(patchwork)
p_celltype<-NULL
for (i in names(spleen_objects)){
  object<-spleen_objects[[i]]
  object$cluster<-as.factor(object$cluster)
  p_celltype[[i]]<-ImageDimPlot(object, fov = i, group.by = "cluster", axes = FALSE, dark.background = F,cols = "glasbey")  + ggtitle(i) +    
      theme(legend.position = "right",
      plot.title = element_text(size = 10),
      plot.margin = margin(0, 0, 0, 0, "null")  # Reduce margins (top, right, bottom, left)
    )
}
wrap_plots(p_celltype,nrow = 3,guides = "collect") & NoLegend()
```

You may take a closer look on the first image

```{r,fig.height=6,fig.width=10}
p_celltype[[1]]
```


Then we add SpaTopic to the Seurat objects for visualization

```{r}
itr_df<-do.call(rbind, dataset)
itr_df$image<-as.factor(itr_df$image)
itr_df$topic<-as.factor(gibbs.res$cell_topics)

## update result in seurat object
for(i in names(spleen_objects)){
    object<-spleen_objects[[i]]
    name2<-gsub("\\.", "-", i)
    object$Topic<-itr_df$topic[itr_df$image == i]
    spleen_objects[[i]]<-object
}
```

We also plot the spatial distribution of the main six topics.

```{r topic, fig.width=12, fig.height=12}
palatte<- c("#0000FFFF","#FF0000FF","#00FF00FF","#009FFFFF","#FF00B6FF","#005300FF","#FFD300FF")
names(palatte)<-as.character(1:7)

p_spatopic<-NULL
for (i in names(spleen_objects)){
  object<-spleen_objects[[i]]
  object$Topic<-as.factor(object$Topic)
  p_spatopic[[i]]<-ImageDimPlot(object, fov = i, group.by = "Topic", axes = FALSE, dark.background = F,cols = palatte)  + ggtitle(i) +    
      theme(legend.position = "right",
      plot.title = element_text(size = 10),
      plot.margin = margin(0, 0, 0, 0, "null")  # Reduce margins (top, right, bottom, left)
    )
}
wrap_plots(p_spatopic,nrow = 3,guides = "collect") & NoLegend()
```

Let's take a closer look on the first image again

```{r,fig.height=6,fig.width=10}
p_spatopic[[1]]
```


We use a heatmap to show the cell type distributions for all six topics:

- Normal spleen samples are primarily characterized by topics 1, 2, and 3, which reflect red pulp (mixed of B cells, erythroblasts, and F4/80(+) macrophages), PALS (mixed of CD8 T cells and CD4 T cells), and B-follicle (B cell dominated) respectively. 
- There is an increase in Topic 6 and depletion of Topic 1 in MRL samples, representing much fewer B cells and F4/80(+) macrophages but more granulocytes and erythroblasts in the red pulp regions. 
- Topic 5 (mainly B220(+) DN T cells and CD4(+) T cells) is enriched in lupus-affected spleen tissue at late stage. 

```{r,fig.height=5,fig.width=10}
m <- as.data.frame(gibbs.res$Beta)

library(RColorBrewer)
breaksList = seq(0.01, 0.4, by = 0.01)

# Plots the heatmap
pheatmap::pheatmap(t(m), 
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdYlBu")))(length(breaksList)), 
         angle_col = 45,
         angle_row = 45,
         breaks = breaksList) # Sets the breaks of the color scale as in breaks List
```
