---
title: "Breast cancer 10x Xenium data analysis"
author: "Nikki Xiao"
date: "2026-01-30"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Set up

We use Seurat package and ggplot2 to visualize the results. For large datasets, we also increase the global memory limit to avoid errors.

```{r}
library(Seurat)
library(ggplot2)
options(future.globals.maxSize = 2 * 1024^3)
```

We use a 10x Genomics Xenium breast cancer dataset to illustrate how to use `SpaTopic`. The data object here can be download from [here](https://cf.10xgenomics.com/samples/xenium/1.0.1/Xenium_FFPE_Human_Breast_Cancer_Rep1/Xenium_FFPE_Human_Breast_Cancer_Rep1_outs.zip), with original public resources available on the [10x Genomics website](https://www.10xgenomics.com/products/xenium-in-situ/preview-dataset-human-breast). We use the Seurat function `ReadXenium()` to load both **centroid** and **segmentation** spatial information from Xenium in-situ data, which will later be used for spatial visualization and topic inference.

```{r}
#### Sample 1 ####
path <- "/Users/nikkixiaomengqi/Downloads/research/outs/"
breastCa.Rep1 <- ReadXenium(
  data.dir = path,
  type = c("centroids", "segmentations"),
)
```

We build spatial coordinate objects using both cell centroids and segmentation boundaries provided by the Xenium output. Centroids represent the cell locations, and segmentation boundaries capture detailed cell shapes. These spatial components are combined into a single **Field of View (FOV)** object using `CreateFOV()` function, ensures that gene expression, cell identity, and spatial coordinates are consistently aligned within the Seurat object.

We then create a Seurat object using the Xenium gene expression matrix. In addition to the gene expression assay, we also store blank codewords and negative controls as separate assays.

```{r}
assay <- "Xenium"
segmentations.data <- list(
  "centroids" = CreateCentroids(breastCa.Rep1$centroids),
  "segmentation" = CreateSegmentation(breastCa.Rep1$segmentations)
)
coords <- CreateFOV(
  coords = segmentations.data,
  type = c("segmentation", "centroids"),
  molecules = breastCa.Rep1$microns,
  assay = assay
)

breastCa.Rep1.obj <- CreateSeuratObject(counts = breastCa.Rep1$matrix[["Gene Expression"]], assay = assay)
breastCa.Rep1.obj[["BlankCodeword"]] <- CreateAssayObject(counts = breastCa.Rep1$matrix[["Blank Codeword"]])
breastCa.Rep1.obj[["ControlCodeword"]] <- CreateAssayObject(counts = breastCa.Rep1$matrix[["Negative Control Codeword"]])
breastCa.Rep1.obj[["ControlProbe"]] <- CreateAssayObject(counts = breastCa.Rep1$matrix[["Negative Control Probe"]])

## We add the spatial coordinate information to the Seurat object under the corresponding FOV for later spatial visualization.
fov <- "Rep1"
breastCa.Rep1.obj[[fov]] <- coords
```

We load cell type annotations for this dataset from a previously curated reference and loaded here for visualization and interpretation.

```{r}
## cell type
anno.R1<-read.csv(file = "/Users/nikkixiaomengqi/Downloads/research/GSM7780153_Xenium_R1_Fig1-5_supervised.csv")
breastCa.Rep1.obj[["Celltype"]]<-anno.R1$Cluster
```

Load the [file]() below to skip the data preparation steps above.

```{r}
load(file = "/Users/nikkixiaomengqi/Downloads/research/breastCa.Rep1.obj.Rdata")
```

We first visualize the spatial distribution of annotated cell types using `ImageDimPlot()`. This provides an overview of tissue organization, and generates a baseline for comparison with SpaTopic-derived spatial topics. In this plot, each point represents a cell, colored by its annotated cell type, overlaid on the spatial coordinates of the tissue section.

```{r}
breastCa.Rep1.obj$Celltype<-as.factor(breastCa.Rep1.obj$Celltype)
celltype.plot <-ImageDimPlot(breastCa.Rep1.obj, group.by = "Celltype", fov = "Rep1", axes = TRUE, cols = "glasbey",dark.background = F,flip_xy = FALSE)+ ggtitle("Cell type")+theme_bw()
celltype.plot
```

## Zoom in

We need to redefine the crop function because the Crop() function behaves as if the x and y coordinates are swapped. To be consistent, we define a small wrapper function that internally switches the x and y inputs.

```{r}
### Redefine the crop function
### x and y seems be swapped in this function x = y, y = x
### need switch x and y to be consistent
#tumor.crop <- Crop(breastCa.Rep1.obj[["Rep1"]], x =  c(2000,4000), y = c(1000,2000))
Crop_custom<-function(objects,x, y){
  Crop(objects, x =  y, y = x)
}
```

We then use this function to define multiple zoomed-in regions from the original image, and then store them as a new FOV within the Seurat object, uses cell segmentation boundaries for visualization.

```{r}
tumor.crop<-Crop_custom(breastCa.Rep1.obj[["Rep1"]], x = c(1000,2000), y =  c(2000,4000))
breastCa.Rep1.obj[["zoom2"]] <- tumor.crop
DefaultBoundary(breastCa.Rep1.obj[["zoom2"]]) <- "segmentation"
```

```{r}
LN.crop<-Crop_custom(breastCa.Rep1.obj[["Rep1"]], x = c(1500,2500), y =  c(2000,4000))
breastCa.Rep1.obj[["zoom1"]] <- LN.crop
DefaultBoundary(breastCa.Rep1.obj[["zoom1"]]) <- "segmentation"
```

Now we visualize the spatial distribution of annotated cell types within the zoom-in region using the function `ImageDimPlot`. This plot shows the spatial patterns that are difficult to observe at the full-image scale.

```{r}
ImageDimPlot(breastCa.Rep1.obj, group.by = "Celltype", fov = "zoom1", axes = TRUE, cols = "glasbey", dark.background = F,flip_xy = FALSE)+ ggtitle("Cell type")+theme_bw()
```

Also the molecule coordinates for selected marker genes within the same zoomed region.

```{r}
ImageDimPlot(breastCa.Rep1.obj, fov = "zoom1", group.by = NA,molecules = c("IL7R", "MS4A1"),dark.background = F, nmols = 20000, alpha = 0.1,mols.size = 0.3,mols.alpha = 1,axes = FALSE,flip_xy = FALSE)
```

You can also download the updated object [here]() and load it to skip these preprocessing steps.

```{r}
save(breastCa.Rep1.obj, file = "breastCa.Rep1.obj.Rdata")
```

How to extract molecule information for each gene:

```{r}
# Get the image object for your FOV
img <- breastCa.Rep1.obj[["zoom2"]]

# Inspect slots
slotNames(img)

# If `molecules` exists, try:
mol_data <- img@molecules
```
